Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
more

@@

\DeclareFontFamily{OT1}{pzc}{}
\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
\DeclareMathAlphabet{\mathcalx}{OT1}{pzc}{m}{it}

@@

\newcommand{\boolop}{\con{b}}
\newcommand{\forallf}[1]{\forall_{\!#1\:}}
\newcommand{\fnull}{\emptyset}
\newcommand{\vecfarg}[1]{\vec{\concret}_{\vthinspace#1}}
\newcommand{\farg}[1]{\concret_{\vthinspace#1}}
\newcommand{\fargx}[2]{\concret^{\vthinspace#2}_{\vthinspace#1}}
\newcommand{\fargsort}[1]{\upsilon_{\vthinspace#1}}

@@

Unfortunately, these fragments cannot accommodate the vast majority of
quantified formulas that correspond to recursive function definitions.
In particular, the fragments in~\cite{GeDeM-CAV-09} are not applicable when the argument of a recursive function is used in various ways (e.g. within an arithmetic expression), 
while the fragment in~\cite{ReyEtAl-1-RR-13} is not applicable for functions over infinite domains such as the integers.
So, for instance, the quantified formula $\forall x : {\sortint}.\;\, \lite( p \leq 0, p( x ) \teq 0, p( x ) \teq 2 * p( x - 1 ) )$ 
is not in the essentially uninterpreted fragment (since $x$ occurs as a child of $-$),
nor does finite model finding apply (since $x$ is of type $\sortint$).

@@

\begin{definition}
Given a set of function definitions $\mathcal{D} = \{ \forallf{f_1} \vec x. \varphi_1, \ldots, \forallf{f_n} \vec x. \varphi_n \}$,
a formula $\varphi$ is \emph{closed under function expansion with respect to $\mathcal{D}$} if
\begin{eqnarray}
\varphi \text{ entails } \displaystyle\bigwedge\limits_{i=1}^n \{ \varphi_i[ \vec t / \vec x ] \mid f_i( \vec t ) \in \terms( \varphi ) \}
\end{eqnarray}
The set $\mathcal{D}$ is \emph{admissible} if whenever there exists a model for $\varphi$ that is closed under function expansion,
there exists a model for $\varphi \wedge \forallf{f_1} \vec x. \varphi_1 \wedge \ldots \wedge \forallf{f_n} \vec x. \varphi_n$.
\end{definition}

For instance, the set $\{ \forallf{f} \vec x. f( x ) \teq 0 \}$ is admissible.
A set of admissible function definitions can include definitions of non-terminating functions, e.g.
$\{ \forallf{f} x : \sortint.\;\, f( x ) \teq f( x - 1 ) \}$ or even $\{ \forallf{f} x. f( x ) \teq f( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{f} x. f( x ) \teq f( x ) + 1 \}$ is not admissible, 
since the empty set of formulas is closed under function expansion with respect to this set,
and there is no model satisfying $\forallf{f} x. f( x ) \teq f( x ) + 1$.
Consider the set $\{ \forallf{f} x. f( x ) \teq f( x ), \forallf{g}. \lite( \neg f( x ) \teq 5, g( x ) \teq g( x ) + 1, g( x ) \teq f( x ) ) \}$.
While this set has a model where $f$ and $g$ are interpreted as the constant $5$, it is not admissible
since the formula $f( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $f( 0 ) \teq 4$ and $\forallf{g}. \lite( \neg f( x ) \teq 5, g( x ) \teq g( x ) + 1, g( x ) \teq f( x ) )$.

@@

Let us fix a signature $\Sigma$, a $\Sigma$-formula
$\varphi$ in definitional form with respect to functions $\sfundefs{\Sigma}$,
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature
$\extendsig{ \Sigma }$ and a $\extendsig{ \Sigma }$-formula $\psi$ such that
$\varphi$ and $\psi$ are equisatisfiable.
\jb{Equisatisfiability as defined in Sect.\ 2 requires the same $\Sigma$ and
the same $T$ for both formulas, which isn't the case here.}
The intuition behind this translation
is to use an uninterpreted type \ct{to be defined} to abstract the set of
\relax{relevant} tuples for each recursive function $\con{f}$ and restrict the
quantification of the function definition for $\con{f}$ to a variable of this
type. Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.

@@

\begin{example}
\label{ex:translation}
Consider the combined signature of linear arithmetic and uninterpreted
functions $\Sigma = \Sigmalia \cup \Sigma_{\euf}$, where
$\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and
$\sfuns{\Sigma_{\euf}} = \{ \con{s} : { \sortint \rightarrow \sortint },\, \const{c} : { \sortint } \}$.
Let $\varphi$ be the $\Sigma$-formula
\begin{equation} \label{eq:ex-before}
\forallf{\con{s}} x : {\sortint}.\;\, \lite( x \leq 0,\, s(x) \teq 0,\, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation} 
%
Here, $s$ is a function that returns the sum of positive numbers between $0$
and its argument $x$. The formula $\varphi$ is in definitional form with
respect to functions $\sfundefs{\Sigma_u}$, and states that the sum of all
positive numbers between $0$ and $c$ is greater than $100$. This formula is
satisfiable with a model that interprets $c$ as an integer greater than or
equal to $14$. Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ includes the type
$\fargsort{s}$, and the uninterpreted function $\farg{s}$ of type $\fargsort{s}
\rightarrow \sortint$. The result of $\conv( \varphi, \ptrue )$ is the pair $(
\psi, \emptyset )$, where after simplification, $\psi$ is the
$\extendsig{\Sigma}$-formula
%
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}.\;\, \lite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargsort{s}$, thus enabling finite model finding
techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for
$\fargsort{s}$. Moreover, all occurrences of $y$ in this formula are beneath
applications of the uninterpreted function $\farg{s}$, implying that the
formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09},
for which there exists a complete instantiation procedure. We observe that both
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), and both
produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second. $\square$
\end{example}

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{equation} \label{eq:approx-interp}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{equation}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\I$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ interprets e.g.\ $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are admissible, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp.\ nonexistence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For example, the model $\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$, 
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with each model of $\conv( \varphi )$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is not an issue, because
$s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.

@@

Let us revisit the translated formula~(\ref{eq:ex-after}) from Example~\ref{ex:translation}.
This formula is unsatisfiable if and only if formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargsort{f}$ can be interpreted as $\sortint$,
and $\farg{f}$ can be interpreted as the identity function.
Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if and only if formula~(\ref{eq:ex-before}) is satisfiable,
noting that the set $\{ \forall_{s} x\typ{\sortint}.\;\, \lite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \}$ is admissble.
As stated in the proof of the previous theorem, 
the reason is that any model of formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under function expansion,
hence implying that formula~(\ref{eq:ex-before}) has a model.
In more detail, let $\varphi_s$ be the formula $\forall_{s} x. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) )$.
The rightmost conjunct 
$\exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c$ ensures 
that any model of formula~(\ref{eq:ex-after}) satisfying $s( c ) > 100$ also satisfies $\varphi_s[c/x]$.
Similarly, the nested existential $\exists z. \farg{s}( z ) \teq \farg{s}( y )-1 )$
ensures that this constraint is enforced recursively.
Given a model $\I$ of formula~(\ref{eq:ex-after}) that interprets $c$ as $14$, 
it is easy to see that $\I$ must satisfy the formula:
\begin{eqnarray} \label{eq:fun-closure}
\begin{split}
\psi := s( c ) > 100 \wedge & \displaystyle\bigwedge\limits_{i=0}^{13} ( \neg (c-i \leq 0) \wedge s( c-i ) \teq i + s( c-i-1 ) ) \\
                            & \wedge c-14 \leq 0 \wedge s( c-14 ) \teq 0
\end{split}
\end{eqnarray}
Notice that $\psi$ is closed under function expansion, 
since it contains the $s$-applications that are equivalent to $s( c-i )$ for $i = 0, \ldots, 14$, and
entails $\varphi_s[c-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_s$ is an admissible definition,
this implies there exists a model for $\psi \wedge \varphi_s$,
which implies there exists a model for formula~(\ref{eq:ex-before}).

@@

When reading this input, 
\cvc adds the (annotated) quantified formula $\forallf{s} x.\;\, s( x ) \teq \lite( x \leq 0, 0, s( x-1 ) )$ to its list of assertions,
which after rewriting becomes $\forallf{s} x.\;\, \lite( x \leq 0, s( x ) \teq 0, s( x ) \teq s( x-1 ) )$.
If \cvc's finite model finding mode for recursive functions is enabled (using the command line parameter ``--fmf-fun"),
then it will replace its list of known assertions based on the conversion $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible, 
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output an approximation of the interpretation for recursive function definitions,
that is, it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.

>>>>

\newcommand\xend{{\hfill$\scriptstyle\blacksquare$}}

@@

%%% for \mathcalx
\DeclareFontFamily{OT1}{pzc}{}
\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
\DeclareMathAlphabet{\mathcalx}{OT1}{pzc}{m}{it}

%%% for correctly sized Greek
\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}

@@

\newcommand{\boolop}{\con{b}}
\newcommand{\forallf}[1]{\forall_{\!#1\:}}
\newcommand{\fnull}{\emptyset}
\newcommand{\vecfarg}[1]{\vec{\concret}_{#1}}
\newcommand{\farg}[1]{\concret_{#1}}
\newcommand{\fargx}[2]{\concret^{\vthinspace#2}_{#1}}
\newcommand{\fargsort}[1]{\alpha_{#1}}

@@

Unfortunately, these fragments cannot accommodate the vast majority of
quantified formulas that correspond to recursive function definitions.
In particular, the fragments in~\cite{GeDeM-CAV-09} are not applicable when the argument of a recursive function is used in various ways (e.g. within an arithmetic expression), 
while the fragment in~\cite{ReyEtAl-1-RR-13} is not applicable for functions over infinite domains such as the integers.
So, for instance, the quantified formula $\forall x : {\sortint}.\;\, \lite\bigl( p \leq 0, p( x ) \teq 0, p( x ) \teq 2 * p( x - 1 ) \bigr)$ 
is not in the essentially uninterpreted fragment (since $x$ occurs as a child of $-$),
nor does finite model finding apply (since $x$ is of type $\sortint$).

@@

\begin{definition}
Given a set of function definitions $\mathcal{D} = \{ \forallf{f_1} \vec x.\; \varphi_1, \ldots, \forallf{f_n} \vec x.\; \varphi_n \}$,
a formula $\varphi$ is \emph{closed under function expansion with respect to $\mathcal{D}$} if
\begin{eqnarray}
\varphi \text{ entails } \displaystyle\bigwedge\limits_{i=1}^n \{ \varphi_i[ \vec t / \vec x ] \mid f_i( \vec t ) \in \terms( \varphi ) \}
\end{eqnarray}
The set $\mathcal{D}$ is \emph{admissible} if whenever there exists a model for $\varphi$ that is closed under function expansion,
there exists a model for $\varphi \wedge \forallf{f_1} \vec x.\; \varphi_1 \wedge \ldots \wedge \forallf{f_n} \vec x.\; \varphi_n$.
\end{definition}

For instance, the set $\{ \forallf{\con{f}} \vec x.\; \con{f}( x ) \teq 0 \}$ is admissible.
A set of admissible function definitions can include definitions of non-terminating functions, e.g.
$\{ \forallf{\con{f}} x : \sortint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not admissible, 
since the empty set of formulas is closed under function expansion with respect to this set,
and there is no model satisfying $\forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1$.
Consider the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ), \forallf{\con{g}}.\;\, \lite\bigl( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) \bigr) \}$.
While this set has a model where $f$ and $g$ are interpreted as the constant $5$, it is not admissible
since the formula $\con{f}( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $\con{f}( 0 ) \teq 4$ and $\forallf{\con{g}}.\;\, \lite( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) )$.

@@

Let us fix a signature $\Sigma$, a $\Sigma$-formula
$\varphi$ in definitional form with respect to $\sfundefs{\Sigma}$,
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature
$\extendsig{ \Sigma }$ and a $\extendsig{ \Sigma }$-formula $\psi$ such that
$\varphi$ and $\psi$ are equisatisfiable.
\jb{Equisatisfiability as defined in Sect.\ 2 requires the same $\Sigma$ and
the same $T$ for both formulas, which isn't the case here.}
The intuition behind this translation
is to use an uninterpreted type \ct{to be defined} to abstract the set of
\relax{relevant} tuples for each recursive function $\con{f}$ and restrict the
quantification of the function definition for $\con{f}$ to a variable of this
type. Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.

@@

\begin{example}
\label{ex:translation}
Consider the combined signature of linear arithmetic and uninterpreted
functions $\Sigma = \Sigmalia \mathrel{\cup} \Sigma_{\euf}$, where
$\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and
$\sfuns{\Sigma_{\euf}} = \{ \con{s} : { \sortint \rightarrow \sortint },\, \const{c} : { \sortint } \}$.
Let $\varphi$ be the $\Sigma$-formula
\begin{equation} \label{eq:ex-before}
\forallf{\con{s}} n : {\sortint}.\;\, \lite\bigl( n \leq 0,\; \con{s}(n) \teq 0,\;
  \con{s}( n ) \teq n + \con{s}( n - 1 ) \bigr) \wedge \con{s}( \con{c} ) > 100
\end{equation} 
%
Here, $\const{s}$ is a function that returns the sum of the positive integers
up to $n$. The formula $\varphi$ is in definitional form with
respect to $\sfundefs{\Sigma_u}$ and states that the sum of all
positive numbers between $0$ and $\const{c}$ is greater than $100$. This formula is
satisfiable with a model that interprets $\const{c}$ as an integer greater than or
equal to $14$. Due to the universal quantifier,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ includes the type
$\fargsort{s}$ and the uninterpreted function $\farg{\con{s}}$ of type $\fargsort{s}
\rightarrow \sortint$. The result of $\conv( \varphi,\, \ptrue )$ is the pair $(
\psi,\, \true )$, where (after simplification) $\psi$ is the
$\extendsig{\Sigma}$-formula
%
\begin{equation} \label{eq:ex-after}
\!\begin{aligned}[t]
  & \bigl(\forall a : \fargsort{s}.\;\, \lite\bigl( \farg{\con{s}}( a ) \leq 0,\;
     \con{s}(\farg{\con{s}}( a )) \teq 0,\;
     \con{s}(\farg{\con{s}}( a )) \teq \farg{\con{s}}( a ) + \con{s}( \farg{\con{s}}( a )-1 ) \bigr) \\[-\jot]
 & {} \wedge \exists z\typ{\fargsort{s}}. \farg{\con{s}}( z ) \teq \farg{\con{s}}( a )-1 ) \\[-\jot]
 & {} \wedge ( \con{s}( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{\con{s}}( z ) \teq c )
\end{aligned}
\end{equation}
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargsort{s}$, thus enabling finite model finding
techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for
$\fargsort{s}$. Moreover, all occurrences of $a$ in this formula are beneath
applications of the uninterpreted function $\farg{\con{s}}$, implying that the
formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09},
for which there exists a complete instantiation procedure. We observe that both
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), and both
produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.\xend
\end{example}

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{\con{s}}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{equation} \label{eq:approx-interp}
\lambda x\typ{Int}. \lite\bigl( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )\bigr)
\end{equation}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\I$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ interprets e.g.\ $\con{s}( -1 )$ and $\con{s}( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are admissible, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $\con{s}( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp.\ nonexistence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For example, the model $\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$, 
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with each model of $\conv( \varphi )$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is not an issue, because
$s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.

@@

Let us revisit the translated formula~(\ref{eq:ex-after}) from Example~\ref{ex:translation}.
This formula is unsatisfiable if and only if formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargsort{f}$ can be interpreted as $\sortint$,
and $\farg{f}$ can be interpreted as the identity function.
Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if and only if formula~(\ref{eq:ex-before}) is satisfiable,
noting that the set $\{ \forallf{\con{s}} x\typ{\sortint}.\;\, \lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated in the proof of the previous theorem, 
the reason is that any model of formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under function expansion,
hence implying that formula~(\ref{eq:ex-before}) has a model.
In more detail, let $\varphi_s$ be the formula $\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr)$.
The rightmost conjunct 
$\exists z\typ{\fargsort{s}}. \farg{\con{s}}( z ) \teq c$ ensures 
that any model of formula~(\ref{eq:ex-after}) satisfying $\con{s}( c ) > 100$ also satisfies $\varphi_s[c/x]$.
Similarly, the nested existential $\exists z. \farg{\con{s}}( z ) \teq \farg{\con{s}}( y )-1 )$
ensures that this constraint is enforced recursively.
Given a model $\I$ of formula~(\ref{eq:ex-after}) that interprets $c$ as $14$, 
it is easy to see that $\I$ must satisfy the formula:
\begin{eqnarray} \label{eq:fun-closure}
\begin{split}
\psi := \con{s}( c ) > 100 \wedge & \displaystyle\bigwedge\limits_{i=0}^{13} ( \neg (c-i \leq 0) \wedge \con{s}( c-i ) \teq i + \con{s}( c-i-1 ) ) \\
                            & \wedge c-14 \leq 0 \wedge \con{s}( c-14 ) \teq 0
\end{split}
\end{eqnarray}
Notice that $\psi$ is closed under function expansion, 
since it contains the $s$-applications that are equivalent to $\con{s}( c-i )$ for $i = 0, \ldots, 14$, and
entails $\varphi_s[c-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_s$ is an admissible definition,
this implies there exists a model for $\psi \wedge \varphi_s$,
which implies there exists a model for formula~(\ref{eq:ex-before}).

@@

When reading this input, 
\cvc adds the (annotated) quantified formula $\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0, 0, \con{s}( x-1 )\bigr)$ to its list of assertions,
which after rewriting becomes $\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0, \con{s}( x ) \teq 0, \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$.
If \cvc's finite model finding mode for recursive functions is enabled (using the command line parameter ``--fmf-fun"),
then it will replace its list of known assertions based on the conversion $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible, 
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output an approximation of the interpretation for recursive function definitions,
that is, it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.
